---
title: "Peering into the twilight - a systematic review on visual-based investigation
  of mesophotic ecosystems - POLAR"
author: "Claudia Campanini"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
  html_document:
    toc: true
    number_sections: true
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, tidy = "styler"
)
```

# DATA PREPARATION

## Load libraries

```{r, show = F}
library(here)
library(dplyr)
library(ggplot2)
library(readxl)   # to import excel data sets
library(magrittr)
library(stringr) #many data wrangling functions
library(readxl)
```

## Load data

```{r}
polar <- read_excel(here("Extracted_data_analysis","Input_data","extracted_raw.xlsx"))
```

## Subset for studies encompassing polar latitudinal region

```{r}
polar <- subset(polar, str_detect(polar$latitudinalRegion, "polar"))
```

### Cleaning

Clean year columns 

```{r Clean columns }
#keep the year only in the year column (in some rows exact time is specified)
polar$publicationYear <-  stringr::str_remove(polar$publicationYear, 
                                 pattern = "\\-.*") 
```

## MISSING / ND / NOTAPP VALUES

Create a summary table 

```{r}
library(purrr)
library(tibble)

# Get all column names
cols <- names(polar[,-c(1:13)])

# Create a summary list
summary_list <- map(cols, function(col) {
  data <- polar[,-c(1:13)][[col]]
  
  tibble(
    Column    = col,
    unique    = paste(unique(data), collapse = ", "),
    nd        = sum(data == "nd", na.rm = TRUE),
    notapp    = sum(data == "notapp", na.rm = TRUE),
    assigned  = sum(!is.na(data) & data != "nd" & data != "notapp"),
    `NA`        = sum(is.na(data)),
    total     = sum(data == "nd", na.rm = TRUE) +
                sum(data == "notapp", na.rm = TRUE) +
                sum(!is.na(data) & data != "nd" & data != "notapp")
  )
})

# Combine into one dataframe
summary_df <- bind_rows(summary_list)

summary_df

rm(summary_list, cols)
```


# GEOGRAPHICAL DISTRIBUTION

```{r}
location <- polar[,c (
  "recordTitle","DOI","publicationYear","prov_code","ecoreg_code")]
```

Load Marine Ecoregion of the World data

```{r Load Marine Ecoregion of the World data}
meow <- read.csv(file=here("Extracted_data_analysis","Input_data","MEOW.csv"))
```

## Provinces

```{r}
prov <- location

prov_lvls <- 
  unique(
    unlist(
      str_split(location$prov_code, "\\_")
                               )
                        )


prov[,prov_lvls] <- sapply(prov_lvls, 
                                      function(y){
                                        str_count(location$prov_code, y)
                                        }
                                      )
rm(prov_lvls)
```

Create a column with the sum of provinces per publication

```{r}
prov %<>%
   mutate(
     prov_per_record = rowSums(
     across(
       where(is.numeric)
       )
     )
   )

prov$prov_per_record <- as.factor(prov$prov_per_record)
```

```{r}
prov_per_record_df <- prov %>% 
  group_by(prov_per_record) %>%
  summarise(n_records = length(prov_per_record)) %>%
  mutate(percentage= round(n_records / sum(n_records) * 100,0))

prov_per_record_df
```

Create a data frame with number of publication per province

```{r}
prov_tot <- 
  prov [, -(length(prov)-1)] %>% # filter out the sum column
  summarise(across(where(is.numeric), 
                   list(sum)
                   )
            )

prov_tot <- reshape2::melt(prov_tot, id.vars = NULL) 

colnames(prov_tot) <- c("Province", "Count")

prov_tot %<>%
      mutate_at("Province", 
                stringr::str_replace, "_1", "")
```

## Ecoregions

```{r create ecoreg}
ecoreg <- location
ecoreg_lvls <- 
  unique(
    unlist(
      str_split(ecoreg$ecoreg_code, "\\_")
                               )
                        )

ecoreg_lvls<- ecoreg_lvls[! ecoreg_lvls %in% c("")]

ecoreg[,ecoreg_lvls] <- sapply(ecoreg_lvls, 
                                      function(y){
                                        str_count(ecoreg$ecoreg_code, y)
                                        }
                                      )
```

Create a column with the number of ecoregions per study

```{r Create a column with the number of ecoregions per study}
ecoreg %<>%
   mutate(
     ecoreg_per_record = rowSums(
     across(
       where(is.numeric)
       )
     )
   )

ecoreg$ecoreg_per_record <- as.factor(ecoreg$ecoreg_per_record)
```

Calculate number of records by number of ecoregions in %

```{r}
ecoreg_per_record_df <- ecoreg %>% 
  group_by(ecoreg_per_record) %>%
  summarise(n_records = length(ecoreg_per_record)) %>%
  mutate(percentage= round(n_records / sum(n_records) * 100,1))

ecoreg_per_record_df
```

Create a data frame with number of publication per ecoregion

```{r Create a data frame with number of publication per ecoregion}
ecoreg_tot <- 
  ecoreg [, -(length(ecoreg)-1)] %>% # filter out the sum column
  summarise(across(where(is.numeric), 
                   list(sum)
                   )
            )

ecoreg_tot <- reshape2::melt(ecoreg_tot, id.vars = NULL) 

colnames(ecoreg_tot) <- c("Ecoregion", "Count")

ecoreg_tot %<>%
      mutate_at("Ecoregion", 
                stringr::str_replace, "_1", "")
```

Add count column to Meow and set 0 to all the ecoregions where no study was conducted

```{r Add count column to Meow and set 0 to all the ecoregions where no study was conducted}

colnames(ecoreg_tot)[colnames(ecoreg_tot) == 'Ecoregion'] <- 'ECO_CODE'
ecoreg_tot$ECO_CODE <- as.character(ecoreg_tot$ECO_CODE)
meow$ECO_CODE <- as.character(meow$ECO_CODE)


ecoreg_tot <- 
  left_join(meow, ecoreg_tot, by = 'ECO_CODE') %>% 
  mutate(Count = replace(Count, is.na(Count), 0))
```

Export count per ecoregion to csv for the map in QGIS

```{r, eval= F}
writexl::write_xlsx(ecoreg_tot,
           here("Extracted_data_analysis","polar","Output-Transformed_data","eco_reg_tot.xlsx"))
```

## Realms

```{r warning = F}
realms <- ecoreg_tot %>%
    group_by(RLM_CODE, REALM) %>%
  summarise(Count=sum(Count))

realms %<>%
  rename(Code = RLM_CODE,
         Realm = REALM)
```

```{r, eval = F}
writexl::write_xlsx(realms, 
           here("Extracted_data_analysis","polar","Output-Transformed_data","Realm_count.xlsx"))
```


## Export geographical distribution summary tables

```{r, eval= F}
writexl::write_xlsx(list(realms, prov_tot, prov_per_record_df, ecoreg_tot, ecoreg_per_record_df),
           here("Extracted_data_analysis","polar","Output-Transformed_data","geographical_distribution_summary.xlsx"))
```

```{r clean latitudinalRegion, echo=FALSE, results='hide'}
rm(ecoreg_lvls, meow, ecoreg, prov, location, ecoreg_per_record_df, prov_tot, prov_per_record_df)
```

# COUNT COLUMNS WITH COMBO

Create a subset of columns

```{r}
count_columns <- polar %>% 
  select(File,
         meso_substrate,
         habit,
         environmentalPosition,
         observationalORexperimental,
         taxonomicTargetLevel,   
         indicators,
         collectedDataType,
         molecularAnalyses)
```

Prepare data set with columns that need count only and add percentage column

```{r , warning=FALSE}

count_columns[,] <- lapply(count_columns[,], as.factor)

count_results <- list()

for (col_name in colnames(count_columns)) {
  counts <- count(count_columns, !!sym(col_name))
  
  count_results[[col_name]] <- counts
  
  colnames(count_results[[col_name]]) <- c(col_name, "Count")
  
  count_results[[col_name]] %<>%
    filter(
      if_all(.cols = everything(), ~ !grepl("notapp", .x))
    ) %>%
    tidyr::drop_na()

  count_results[[col_name]] %<>%
    mutate(Percentage= round(Count / sum(Count) * 100,2))
}

rm(col_name, counts)

count_results <- count_results[names(count_results) != "File"]
```

### Export summary tables

```{r export summary_counts_question_1-3 }
library(openxlsx)

write.xlsx(count_results,
           file = here("Extracted_data_analysis",
                       "polar",
                       "Output-Transformed_data",
                    "summary_counts_question_1-3.xlsx"))
```

Create treemaps charts 

```{r treemap_counts}
treemap_func = function(dat,var) {
  ggplot(dat,
       aes(area=Count,
           fill=.data[[var]],
           line = "black",
           label=paste0(.data[[var]], 
                        " \n ",Percentage,"%"))
       )+ 
  treemapify::geom_treemap(layout = "squarified")+ 
  treemapify::geom_treemap_text(place = "centre",
                    size = 20)+
  labs(title = var)+
    theme(plot.title = element_text(hjust = 0.5, size= 20),
          legend.position = "none")
  
}

treemap_counts <- list()

for (col_name in colnames(count_columns[,-1])) {
  
  treemap_counts[[col_name]] <- treemap_func(dat= count_results[[col_name]], var = col_name)

}

treemap_counts[[1]]
```

Save all tree maps plots

```{r export treemap_counts}
lapply(names(treemap_counts), function(x){
  ggsave(filename=paste0(x,".tiff"), 
         plot=treemap_counts[[x]],
         units = "in", width=12, height=4, dpi=300, compression = 'lzw',
         path = here("Extracted_data_analysis",
                       "polar",
                       "Output-Visualization", 
                     "Treemap_counts"),
         create.dir = T
         )
  })

lapply(names(treemap_counts), function(x){
  ggsave(filename=paste0(x,".png"), 
         plot=treemap_counts[[x]],
         units = "in", width=12, height=4, dpi=300,
         path = here("Extracted_data_analysis",
                       "polar",
                       "Output-Visualization", 
                     "Treemap_counts"),
         create.dir = T
         )
  })
```

# COUNT METHODS (PURELY TAXONOMIC STUDIES EXCLUDED)

Remove purely taxonomic studies and subset columns
(taxa++ was assigned to studies that led to the description of new species but where also ecological e.g. describing community for a given taxonomic group)

```{r}
methods <- polar %>% 
  filter(taxonomic_publication == "taxa++" |  is.na(taxonomic_publication) ) %>%
  select(File,
         meso_design,
         meso_replicates,
         temporalReplicates,
         abioticVariablesCollection,
         abioticVariables)

count_columns_method <- methods
```

Prepare data set with columns that need count only and add percentage column

```{r, warning=FALSE}
count_columns_method[,] <- lapply(count_columns_method[,], as.factor)

count_results_method <- list()

for (col_name in colnames(count_columns_method)) {
  counts <- count(count_columns_method, !!sym(col_name))
  
  count_results_method[[col_name]] <- counts
  
  colnames(count_results_method[[col_name]]) <- c(col_name, "Count")
  
  # remove rows containing "notapp"
  count_results_method[[col_name]] %<>%
    filter(
      if_all(.cols = everything(), ~ !grepl("notapp", .x))
    ) %>%
    tidyr::drop_na()


  count_results_method[[col_name]] %<>%
    mutate(Percentage= round(Count / sum(Count) * 100,1))
}

rm(col_name, counts)

count_results_method <- count_results_method[names(count_results_method) != "File"]
```

Export summary tables

```{r export summary_counts_question_4.xlsx}
library(openxlsx)

write.xlsx(count_results_method,
           file = here("Extracted_data_analysis", "polar", "Output-Transformed_data",
                    "summary_counts_question_4.xlsx"))
```

Create treemaps charts 

```{r treemap_counts_methods}
treemap_func = function(dat,var) {
  ggplot(dat,
       aes(area=Count,
           fill=.data[[var]],
           line = "black",
           label=paste0(.data[[var]], 
                        " \n ",Percentage,"%"))
       )+ 
  treemapify::geom_treemap(layout = "squarified")+ 
  treemapify::geom_treemap_text(place = "centre",
                    size = 20)+
  labs(title = var)+
    theme(plot.title = element_text(hjust = 0.5, size= 20),
          legend.position = "none")
  
}

treemap_counts <- list()

for (col_name in colnames(count_columns_method[,-1])) {
  
  treemap_counts[[col_name]] <- treemap_func(dat= count_results_method[[col_name]], var = col_name)

}

treemap_counts[[1]]
```

Save all tree maps plots

```{r export tree maps methods, results = 'hide'}
lapply(names(treemap_counts), function(x){
  ggsave(filename=paste0("treemap",x,".tiff"), 
         plot=treemap_counts[[x]],
         units = "in", width=12, height=4, dpi=300, compression = 'lzw',
         path = here("Extracted_data_analysis","polar",                        "Output-Visualization","Treemaps_methods"),
         create.dir = T
         )
  })

lapply(names(treemap_counts), function(x){
  ggsave(filename=paste0("treemap",x,".png"), 
         plot=treemap_counts[[x]],
         units = "in", width=12, height=4, dpi=300,
         path = here("Extracted_data_analysis", "polar", "Output-Visualization","Treemaps_methods"),
         create.dir = T
         )
  })
```

# COUNT COLUMNS SPLIT

Create a list of data frame splitting values

```{r}
split_columns <- polar %>% 
  select(meso_substrate,
         habit,
         environmentalPosition,
         observationalORexperimental,
         taxonomicTargetLevel, 
         indicators,
         collectedDataType,
         molecularAnalyses)


split_columns[,] <- lapply(split_columns[,], as.factor)
```

Prepare data set with columns 

```{r}
split_results <- list()

for (col_name in colnames(split_columns)) {
  # Split the values of each column by "+"
  lvls <- unique(unlist(str_split(split_columns[[col_name]], "\\+")))

  # Initialize a list to store counts for each level
  split_counts <- sapply(lvls, function(y) {
    # Count how many times each unique level appears in the column
    str_count(split_columns[[col_name]], fixed(y))
    }
    )
  
  split_results[[col_name]] <- split_counts
}

rm(lvls)
```

Add percentage column

```{r}
split_tot <- list()

for(col_name in colnames(split_columns)) {
  split_results[[col_name]] <- as.data.frame(split_results[[col_name]])
  
  split_results[[col_name]] %<>% # delete rows with missing values
    select(where(function(x) any(!is.na(x)))) %>%
    select(-any_of("notapp")) %>%
    tidyr::drop_na()
  
    split_tot[[col_name]] <- as.data.frame(split_results[[col_name]]) %>%
    summarise(across(where(is.numeric), list(sum)))

  split_tot[[col_name]]  <- reshape2::melt(split_tot[[col_name]], id.vars = NULL) 

  colnames(split_tot[[col_name]]) <- c(col_name, "Count")

  split_tot[[col_name]] %<>%
        mutate_at(col_name, 
                  stringr::str_replace, "_1", "")
  
    
  split_tot[[col_name]] %<>%
      mutate(Percentage= round(Count / nrow(split_results[[col_name]] ) * 100,2))    
    
}

rm(col_name)
```

Export summary tables

```{r export summary_split_counts_question_1_3.xlsx}
library(openxlsx)

write.xlsx(split_tot,
           file = here("Extracted_data_analysis",                        "polar",                        "Output-Transformed_data",
                    "summary_split_counts_question_1_3.xlsx"))
```

Create bar charts for each collected parameter

```{r barchart_counts}
barchart_func = function(dat,var) {
  
 ggplot(dat, 
       aes(x = reorder(.data[[var]], Count), y = Count, fill= .data[[var]])) +
  geom_bar(stat = "identity", fill = "#0097A7") +
  geom_text(
    data = dat %>% filter(Percentage >= 2),
    aes(label = paste0(round(Percentage, 0), "%")),
    position = position_stack(vjust =0.5),
    color = "black",
    size = 2.5
  )+
  labs(y = "# Publications", x = var) +
  theme_minimal() +
  coord_flip()+  # optional: flips for horizontal bars
  theme(legend.position ="none")
  
}

barchart_counts <- list()

for (col_name in colnames(split_columns)) {
  
  barchart_counts[[col_name]] <- barchart_func(dat= split_tot[[col_name]], var = as.character(col_name))

}

barchart_counts[[1]]
```

```{r}
lapply(names(barchart_counts), function(x){
  ggsave(filename=paste0("bar_",x,".tiff"), 
         plot=barchart_counts[[x]],
         units = "in", width=12, height=4, dpi=300, compression = 'lzw',
         path = here("Extracted_data_analysis",                        "polar",                        "Output-Visualization","Barcharts_split"),
         create.dir = T
         )
  })

lapply(names(barchart_counts), function(x){
  ggsave(filename=paste0("bar_",x,".png"), 
         plot=barchart_counts[[x]],
         units = "in", width=12, height=4, dpi=300,
         path = here("Extracted_data_analysis",                        "polar",                        "Output-Visualization","Barcharts_split"),
         create.dir = T
         )
  })
```

# COUNT METHODS (PURELY TAXONOMIC STUDIES EXCLUDED)

Create data frame with subset

```{r}
split_columns_methods <- methods

split_columns_methods[,] <- lapply(split_columns_methods[,], as.factor)
```

Prepare data set with columns 

```{r}
split_results_methods <- list()

for (col_name in colnames(split_columns_methods)) {
  # Split the values of each column by "+"
  lvls <- unique(unlist(str_split(split_columns_methods[[col_name]], "\\+")))

  # Initialize a list to store counts for each level
  split_counts_methods <- sapply(lvls, function(y) {
    # Count how many times each unique level appears in the column
    str_count(split_columns_methods[[col_name]], fixed(y))
    }
    )
  
  split_results_methods[[col_name]] <- split_counts_methods
}

rm(lvls)
```

Add percentage column

```{r}
split_tot_methods <- list()

for(col_name in colnames(split_columns_methods)) {
  split_results_methods[[col_name]] <- as.data.frame(split_results_methods[[col_name]])
  
  split_results_methods[[col_name]] %<>% # delete rows with missing values
    select(where(function(x) any(!is.na(x)))) %>%
    select(-any_of("notapp")) %>%
    tidyr::drop_na()
  
    split_tot_methods[[col_name]] <- as.data.frame(split_results_methods[[col_name]]) %>%
    summarise(across(where(is.numeric), list(sum)))

  split_tot_methods[[col_name]]  <- reshape2::melt(split_tot_methods[[col_name]], id.vars = NULL) 

  colnames(split_tot_methods[[col_name]]) <- c(col_name, "Count")

  split_tot_methods[[col_name]] %<>%
        mutate_at(col_name, 
                  stringr::str_replace, "_1", "")
  
    
  split_tot_methods[[col_name]] %<>%
      mutate(Percentage= round(Count / nrow(split_results_methods[[col_name]] ) * 100,2))    
    
}

rm(col_name)
```

Export summary tables

```{r export summary_split_counts_question_4.xlsx}
library(openxlsx)

write.xlsx(split_results_methods,
           file = here("Extracted_data_analysis",                        "polar",                        "Output-Transformed_data",
                    "summary_split_counts_question_4.xlsx"))
```

Create bar charts per each parameter collected

```{r barchart_split_methods}
barchart_func = function(dat,var) {
  
 ggplot(dat, 
       aes(x = reorder(.data[[var]], Count), y = Count, fill= .data[[var]])) +
  geom_bar(stat = "identity", fill = "#0097A7") +
  geom_text(
    data = dat %>% filter(Percentage >= 2),
    aes(label = paste0(round(Percentage, 0), "%")),
    position = position_stack(vjust =0.5),
    color = "black",
    size = 2.5
  )+
  labs(y = "# Publications", x = var) +
  theme_minimal() +
  coord_flip()+  # optional: flips for horizontal bars
  theme(legend.position ="none")
  
}

barchart_split_methods <- list()

for (col_name in colnames(split_columns_methods)) {
  
  barchart_split_methods[[col_name]] <- barchart_func(dat= split_tot_methods[[col_name]], var = as.character(col_name))

}

barchart_split_methods[[4]] 
```

Export bar charts in a dedicated folder

```{r export Barcharts_split_methods, results= "hide"}
lapply(names(barchart_counts), function(x){
  ggsave(filename=paste0("bar_",x,".tiff"), 
         plot=barchart_counts[[x]],
         units = "in", width=12, height=4, dpi=300, compression = 'lzw',
         path = here("Extracted_data_analysis",                        "polar",                        "Output-Visualization","Barcharts_split_methods"), create.dir = T
         )
  })

lapply(names(barchart_counts), function(x){
  ggsave(filename=paste0("bar_",x,".png"), 
         plot=barchart_counts[[x]],
         units = "in", width=12, height=4, dpi=300,
         path = here("Extracted_data_analysis",                        "polar",                        "Output-Visualization","Barcharts_split"), create.dir = T
         )
  })
```

# PUBLICATIONS PER DEPTH

```{r}
polar$mininumSurveyedDepthInMeters <- as.numeric(polar$mininumSurveyedDepthInMeters)
polar$maximumSurveyedDepthInMeters <- as.numeric(polar$maximumSurveyedDepthInMeters)

# Define depth bins
bins <- seq(31, 151, by = 10)

# Create a data frame to hold bin counts
pubs_depth <- data.frame(
  bin_start = bins[-length(bins)],
  bin_end = bins[-1]-1,
  count = 0
)

# For each bin, count how many studies cover that depth range
for (i in 1:nrow(pubs_depth)) {
  bin_start <- pubs_depth$bin_start[i]
  bin_end <- pubs_depth$bin_end[i]
  
  pubs_depth$count[i] <- sum(polar$maximumSurveyedDepthInMeters >= bin_start & polar$mininumSurveyedDepthInMeters < bin_end, na.rm = TRUE)
}

pubs_depth$depth_bin <- paste(pubs_depth$bin_start, "-", pubs_depth$bin_end)

pubs_depth$depth_bin <- reorder(pubs_depth$depth_bin, -pubs_depth$bin_start)


ggplot(pubs_depth, aes(x = factor(depth_bin), y = count)) +
  geom_bar(stat = "identity", fill = "#2a788e") +
  labs(x = "Depth interval (m)", y = "Number of publications") +
  theme_minimal()+
  coord_flip()+
  scale_y_continuous(position="right")
```

Export pubs per depth bar plot

```{r Export pubs per depth bar plot, echo = F} 
ggsave("pubs_depth.png", 
       units = "in", width = 8, height = 6,
       path = here ("Extracted_data_analysis",                        "polar",                        "Output-Visualization"))
```

# PLATFORM BY DEPTH

Create a platform by depth data frame and keep meso_samplingPlatform

```{r Create a platform by depth data frame and keep meso_samplingPlatform}

platform_depth <- polar %>%
  select(File, 
         mininumSurveyedDepthInMeters,
         maximumSurveyedDepthInMeters,
         meso_direct_indirect,
         meso_samplingPlatform,
         max_depth_hov, min_depth_hov, 
         max_depth_rov, min_depth_rov,
         max_depth_ccr, min_depth_ccr,
         max_depth_open, min_depth_open,
         max_depth_scuba, min_depth_scuba
         )
```

Split the platform column in one binary column per platform 

```{r Split the platform column in one binary column per platform  }
platform_lvls <- 
  unique(
    unlist(
      str_split(platform_depth$meso_samplingPlatform, "\\+")
                               )
                        )

platform_lvls<- platform_lvls[! platform_lvls %in% c("",NA,"notapp",",","nd")]

platform_depth[,platform_lvls] <- sapply(platform_lvls, 
                                      function(y){
                                        str_count(platform_depth$meso_samplingPlatform, y)
                                        }
                                      )

rm(platform_lvls)

platform_depth[,-c(1:15)] <- lapply(platform_depth[,-c(1:15)], as.numeric)

platform_depth[,c(1:15)] <- lapply(platform_depth[,c(1:15)], as.character)


platform_depth %<>%
  rename(open_direct = open)  %>%
  select(-meso_samplingPlatform)
```

Create a summary table with # studies per platform 

```{r Create a summary table with # studies per platform }
platform_summary <- platform_depth %>%
  summarise(across(where(is.numeric), ~ sum(.x, na.rm = TRUE)))

platform_summary <- as.data.frame(t(platform_summary))

platform_summary %<>%
  mutate(Platform = rownames(platform_summary)) %>%
  rename(Count = V1) %>%
  relocate(Count,.after = where(is.character))

platform_summary
```

Create a column with the number of platforms per study

```{r Create a column with the number of platforms per study}
platform_per_record <- platform_depth %>%
  filter(meso_direct_indirect != "nd") %>%
   mutate(
     platform_per_record = rowSums(
     across(
       where(is.numeric)
       )
     )
   )

platform_per_record$platform_per_record <- as.factor(platform_per_record$platform_per_record)
```

Calculate number of records by number of platforms in %

```{r Calculate number of records by number of platforms in %}
platform_per_record <- platform_per_record %>% 
  group_by(platform_per_record) %>%
  summarise(n_records = length(platform_per_record)) %>%
  mutate(percentage= round(n_records / sum(n_records) * 100,1))

platform_per_record
```

Keep most common visual platforms only

```{r }
platform_depth_mod <- cbind(platform_depth[,c(1:14)],
                            platform_depth$open_direct, 
                            platform_depth$hov, 
                            platform_depth$rov)

names(platform_depth_mod) <- gsub("platform_depth\\$", "", names(platform_depth_mod))
```

Set max depth of studies with max depth > 150 to 150 so they fit into the plot

```{r set max_depth to mesophotic limits, warning= F}
platform_depth_long <- within(platform_depth_mod, rm(mininumSurveyedDepthInMeters, maximumSurveyedDepthInMeters, max_depth_direct, meso_direct_indirect, hov, rov, open_direct, ccr_direct, scuba_direct))

platform_depth_long[,-1] <- lapply(platform_depth_long[,-1], as.numeric)

str(platform_depth_long)

platform_depth_long  %<>%
  mutate(max_depth_scuba = case_when(max_depth_scuba >= 150 ~ 150,
                               TRUE ~ max_depth_scuba)
         ) %>%
  mutate(max_depth_hov = case_when(max_depth_hov >= 150 ~ 150,
                               TRUE ~ max_depth_hov)
         ) %>%
  mutate(max_depth_rov = case_when(max_depth_rov >= 150 ~ 150,
                               TRUE ~ max_depth_rov)
         ) %>%
  mutate(max_depth_ccr = case_when(max_depth_ccr >= 150 ~ 150,
                               TRUE ~ max_depth_ccr)) %>%
  mutate(max_depth_open = case_when(max_depth_open >= 150 ~ 150,
                               TRUE ~ max_depth_open))         
```

Set min depth of studies with min depth <30 to 30 so they fit into the plot

```{r set min_depth to mesophotic limits}
platform_depth_long %<>%
  mutate(min_depth_scuba = case_when(min_depth_scuba <= 30 ~ 30,
                               TRUE ~ min_depth_scuba)
         ) %>%
  mutate(min_depth_hov = case_when(min_depth_hov <= 30 ~ 30,
                               TRUE ~ min_depth_hov)
         ) %>%
  mutate(min_depth_rov = case_when(min_depth_rov <= 30 ~ 30,
                               TRUE ~ min_depth_rov)
         ) %>%
  mutate(min_depth_ccr = case_when(min_depth_ccr <= 30 ~ 30,
                               TRUE ~ min_depth_ccr)) %>%
  mutate(min_depth_open = case_when(min_depth_open <= 30 ~ 30,
                               TRUE ~ min_depth_open))         
```

Convert platform_depth_long to long format

```{r}
platform_min_depth_long <- platform_depth_long %>%
  select(-c(max_depth_hov, max_depth_rov, max_depth_open, max_depth_scuba, max_depth_ccr)) 


platform_min_depth_long[,c(1,2)] <- lapply(platform_min_depth_long[,c(1,2)], as.factor)

platform_min_depth_long %<>% 
  reshape2::melt(platform_min_depth_long, 
                 id.vars = c("File"),
                 measure.vars = grep("^min_depth_|^max_depth_",
                                     names(platform_min_depth_long), 
                                     value = TRUE),
                 variable.name = "Platform",
                 value.name = "min_depth"
                 )
 
platform_max_depth_long <- platform_depth_long %>%
    select(-c(min_depth_hov, min_depth_rov, min_depth_open, min_depth_scuba, min_depth_ccr)) 

platform_max_depth_long %<>%
    reshape2::melt(platform_max_depth_long, 
                 id.vars = c("File"),
                 measure.vars = grep("^min_depth_|^max_depth_",
                                     names(platform_max_depth_long), 
                                     value = TRUE),
                 variable.name = "Platform",
                 value.name = "max_depth"
                 )

platform_depth_lat_long <- cbind(platform_min_depth_long, platform_max_depth_long$max_depth)

rm(platform_min_depth_long, platform_max_depth_long)

platform_depth_lat_long
```


```{r}
platform_depth_lat_long %<>% 
  rename(max_depth =`platform_max_depth_long$max_depth`)

platform_depth_lat_long$Platform <- gsub("min_depth_", "", platform_depth_lat_long$Platform)

platform_depth_lat_long$Platform <- gsub("max_depth_", "", platform_depth_lat_long$Platform)
```

Set proper labelling for platform

```{r}
platform_depth_lat_long %<>%
  mutate(Platform = case_when(Platform == "open" ~ "OC",
                              Platform == "rov" ~ "ROV",
                              Platform == "hov" ~ "HOV",
                              TRUE ~ Platform)
         ) %>%
  mutate(Platform = as.factor(Platform))
```

Add visual column to reorder platform

```{r Add visual column to reorder platform}
platform_depth_lat_long %<>%
  mutate(Visual = as.factor(case_when(Platform == "OC" |
                                      Platform == "SCUBA" |
                                      Platform == "CCR" ~ "direct",
                            TRUE ~ "indirect" )
  )
  ) 

platform_depth_lat_long %<>%
  mutate(Platform = forcats::fct_reorder(Platform, as.integer(Visual)))
```

Create a bar plot with depth bins by platform and latitudinalRegion

```{r  platform_depth_plot barplot}

depth_bins <- tibble(
  bin_start = seq(-30, -140, by = -10),
  bin_end   = seq(-40, -150, by = -10)
) %>%
  mutate(bin_label = paste0(abs(bin_start), "-", abs(bin_end)))

platform_depth_lat_long <- platform_depth_lat_long %>%
  mutate(
    min_depth = as.numeric(min_depth),  # positive e.g., 30
    max_depth = as.numeric(max_depth)   # positive e.g., 150
  )

expanded_data <- platform_depth_lat_long %>%
  tidyr::crossing(depth_bins) %>%
  # Convert study range to negative depths: [-max_depth, -min_depth]
  filter(bin_end >= -max_depth, bin_start <= -min_depth) %>%
  mutate(
    bin_label = factor(bin_label, levels = rev(depth_bins$bin_label))
  )

# Plot with count labels on the right
platform_depth_plot <- ggplot(expanded_data, aes(x = bin_label)) +
  geom_bar(position = "stack", alpha= 0.8, fill= "#2a788e") +
  scale_y_continuous(name = "Number of Studies",
                     expand = expansion(mult = c(0, 0.15))) +  # Make room for text
  scale_x_discrete(name = "Depth (m)") +
  coord_flip() +  # Depth on y-axis (shallow at top)
  facet_wrap(~ Platform, nrow = 1) +
  theme_minimal() +
  theme(legend.position = "none")

platform_depth_plot

```

Export bar plot of most common platform per depth interval

```{r Export bar plot of most common platform per depth interval by latitudinalRegion}

ggsave("platform_depth_plot.tiff", 
       plot = platform_depth_plot,
       units = "in", width = 8, height = 6, dpi = 300, compression = 'lzw',
       path = here ("Extracted_data_analysis",                        "polar",                        "Output-Visualization"))

ggsave("platform_depth_plot.png",
       plot = platform_depth_plot,
       units = "in", width = 8, height = 5,
       path = here ("Extracted_data_analysis", "polar",                      "Output-Visualization"))
```

# DIRECT - INDIRECT BY DECADE

```{r dir_ind_by_decade}
dir_ind_by_decade <- polar

dir_ind_by_decade %<>%
         mutate(decade = case_when(publicationYear >= 1960 & publicationYear < 1970 ~ "1970-1979",
                                 publicationYear >= 1970 & publicationYear < 1980 ~ "1970-1979",
                                 publicationYear >= 1980 & publicationYear < 1990 ~ "1980-1989",
                                 publicationYear >= 1990 & publicationYear < 2000 ~ "1990-1999",
                                 publicationYear >= 2000 & publicationYear < 2010 ~ "2000-2009",
                                 publicationYear >= 2010 & publicationYear < 2020 ~ "2010-2019",
                                 publicationYear >= 2020 ~ "2020-2023")
              ) 

dir_ind_by_decade_lat <- dir_ind_by_decade

dir_ind_by_decade <- dir_ind_by_decade [, c("meso_direct_indirect","decade")] %>%
  group_by(decade,meso_direct_indirect) %>% 
  tally()

colnames(dir_ind_by_decade) <- c("Decade", "Technique","Count")

unique(dir_ind_by_decade$Technique)

dir_ind_by_decade %<>% 
  filter(Technique != "nd" & Technique != "NA") %>% #remove nd from viz because they accounted for a really small percentage
  group_by(Decade) %>%
  mutate(Percentage = round(Count / sum(Count) * 100,1))
```

Make bar plot of technique by decade

```{r Make bar plot of technique by decade}
technique_decade_barplot <- ggplot(subset(dir_ind_by_decade,
              dir_ind_by_decade$Decade != "1970-1979"),
       aes(x=Technique, y=Count, label = Technique)
       ) +
  geom_bar(position="dodge", 
           stat="identity", alpha= 0.8, fill = "#2a788e" ) +
  geom_text(aes(label = paste(round(Percentage,0),"%", sep="")),
            parse = F,
            vjust = -0.25,
            size=3.9)+
  facet_grid(~ Decade, 
             switch = "y") +
  theme_classic() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 10),
        axis.text.y = element_text(size=8.7),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size=13),
        strip.text = element_text(size = 13))+
  labs(y= "Number of publications")

technique_decade_barplot
```

Export bar plot of technique by decade

```{r Export bar plot of technique by decade}
ggsave("technique_decade_barplot.tiff", 
       plot = technique_decade_barplot,
       units = "in", width = 7, height = 6, dpi = 300, compression = 'lzw',
       path = here ("Extracted_data_analysis",                        "polar",                        "Output-Visualization"))

ggsave("technique_decade_barplot.png",
       plot = technique_decade_barplot,
       units = "in", width = 7, height = 6,
       path = here ("Extracted_data_analysis",                        "polar",                        "Output-Visualization"))
```

# FIELD & INDICATORS

Create subset for field & indicators

```{r Clean field_merged data frame }
field_merged <- polar[,c(59:length(polar))]

field_merged[,] <- lapply(field_merged[,], as.numeric)

colSums(field_merged)
```
Select only the most common fields

```{r Select only the most common fields }
field_merged <- field_merged[, colSums(field_merged) > 3]

field_merged$field_Taxonomy <- NULL

colSums(field_merged)


field_merged <- field_merged[, order(colSums(field_merged), decreasing = F)]

colSums(field_merged)
```

Rename fields with long version names

```{r Rename fields with long version names}
field_merged %<>%
  rename(`Biodiversity and \n community structure (17)` = field_commdesc,
         `Distribution and \n connectivity (4)` = field_dist,
         `Disturbances (5)` = field_impass,
         `Trophic ecology (4)` = field_troph,
         `Functional ecology (5)` =  field_func,
         `Mapping (11)` = field_map,
         `Behaviour (5)` = field_Behaviour ,
         `Acoustics (5)` = field_Acoustics ,
         `Environmental drivers and \n spatio-temporal patterns (13)` = field_environmental_drivers
         )
colSums(field_merged)
```

Prepare the indicator matrix 

```{r}
ind_matrix <- split_results[["indicators"]]
```

Select only the most common indicators

```{r Select only the most common indicators}
colSums(ind_matrix)

ind_matrix <- ind_matrix[, order(colSums(ind_matrix), decreasing = TRUE)]

ind_matrix <- as.data.frame(ind_matrix)

ind_matrix <- ind_matrix[!purrr::map_lgl(ind_matrix, ~ all(is.na(.)))] # remove empty column
```

Remove taxonomy studies which have a dedicated section

```{r}
ind_matrix$taxo <- NULL
field_merged$field_taxo <- NULL

colSums(ind_matrix)
```

Rename indicators with long version names

```{r}
ind_matrix %<>%
  rename(`Abundance \n (13)` = Abundance,
         `Cover \n (8)` = Cover,
         `Diversity \n (18)` = Diversity,
         `Habitat \n (9)` = Habitat,
         `Morphology \n (7)` = Morphological,
         `Size \n (4) ` = Size,
         `Recruitment \n (1) ` = Recruitment,
         `Density \n (10)` = Density,
         `Physiological \n indicator \n (1)` = Physiological,
         `Litter \n (1)` = Litter,
         `Biomass \n (4)` = Biomass,
         `Genetic \n sequence \n (2)` = geneticSequences,
         `Functional indicators \n (1)` = func,
         `Growth \n (1)` = growth,
         `Trophic \n (3)` = Trophic,
         `Behaviour \n (4)` = Behaviour
         )

colSums(ind_matrix)
```

Multiply the transpose of the field matrix with the indicator matrix

```{r Multiply the transpose of the field matrix with the indicator matrix}
co_occurrence_matrix <- t(as.matrix(field_merged)) %*% as.matrix(ind_matrix)
```

Reshape the co-occurrence matrix for plotting}

```{r reshape the co-occurrence matrix for plotting}
co_occurrence_df <- reshape2::melt(co_occurrence_matrix)

colnames(co_occurrence_df) <- c("Field", "Indicator", "Count")
```

Plot the heatmap

```{r Plot the heatmap}
field_ind_heat <- ggplot(co_occurrence_df, aes(x = Indicator, y = Field, fill = Count)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "deepskyblue4") +
  geom_text(aes(label = Count), color = "black") +
  theme_minimal()+
  scale_x_discrete(position="top")

field_ind_heat
```

Export plot 

```{r export field_ind_heat}
ggsave("field_ind_heat.png", 
       units = "in", width = 12.5, height = 5.5,
       path = here ("Extracted_data_analysis", "polar",                        "Output-Visualization"))
```

## Calculate # of ecological indicators per record

Create a column with the sum of ecological indicators values to see how many different sampling units were adopted by each study

```{r ecological indicicators per record}
ind_summary <- as.data.frame(split_results[["indicators"]])

ind_summary <- ind_summary[!purrr::map_lgl(ind_summary, ~ all(is.na(.)))] # remove empty column

ind_summary %<>%  rowwise() %>%
   mutate(ind_per_record =  sum(c_across(where(is.numeric))))
```

Calculate number of records by number of sampling units in %

```{r}
ind_summary$ind_per_record <- as.factor(ind_summary$ind_per_record)

ind_per_record <- ind_summary %>% 
  group_by(ind_per_record) %>%
  summarise(n_records = length(ind_per_record)) %>%
  mutate(percentage= round(n_records / sum(n_records) * 100,1))

ind_per_record
```

# SAMPLING UNIT

## Prepare a data frame for sampling unit

```{r Prepare a data frame for sampling unit, results = "hide"}
sampling_df <- polar %>%
  filter(is.na(taxonomic_publication)) %>% # remove purely taxonomic studies
  select(File,recordTitle,DOI,publicationYear,mininumSurveyedDepthInMeters,maximumSurveyedDepthInMeters,samplingUnit)  

sampling_df %>% tidyr::drop_na(samplingUnit)
sampling_df[!is.na(sampling_df$samplingUnit),]

sampling_df %<>% 
  filter(samplingUnit != "notapp")

sampling_df$samplingUnit <- as.character(sampling_df$samplingUnit)
```

Prepare columns per sampling units

```{r prepare column per sampling units}
sampling_lvls <- unique(unlist(str_split(sampling_df$samplingUnit, 
                                         "\\+")
                               )
                        )

rownames(is.na(sampling_df$samplingUnit))

sampling_df[,sampling_lvls] <- sapply(sampling_lvls, 
                                      function(y){
                                        str_count(sampling_df$samplingUnit, y)
                                        }
                                      )

rm(sampling_lvls)

colnames(sampling_df)
```

## Calculate # of sampling unit per record

Create a column with the sum of sampling units values to see how many different sampling units were adopted by each study

```{r sampling units per record}
sampling_df %<>%
   mutate(type_per_record = rowSums(across(quadrat:expunit)
                        )
   )
```

Calculate number of records by number of sampling units in %

```{r}
sampling_df$type_per_record <- as.factor(sampling_df$type_per_record)

sampling_unit_per_record <- sampling_df %>% 
  group_by(type_per_record) %>%
  summarise(n_records = length(type_per_record)) %>%
  mutate(percentage= round(n_records / sum(n_records) * 100,0))

sampling_unit_per_record
```

Create a data frame with number of publication per sampling unit type

```{r Create a data frame with number of publication per sampling unit type}
sampling_unit_tot <- sampling_df  %>%
  summarise(across(c(8:(length(sampling_df)-1)), 
                   list(sum)))

sampling_unit_tot <- reshape2::melt(sampling_unit_tot) 

colnames(sampling_unit_tot) <- c("Sampling_unit", "Count")

sampling_unit_tot %<>%
        mutate_at("Sampling_unit", 
                  stringr::str_replace, "_1", "") 
```

## Simplified sampling units

Create a column with a simplified sampling units for transect and quadrat

```{r Create a column with a simplified sampling units for transect and quadrat}
  sampling_unit_tot %<>%
       mutate(Sampling_unit_macro = case_when(Sampling_unit == 'quadrat' | 
                                  Sampling_unit == "permquad" ~ "Quadrat",
                                  Sampling_unit == "transect"| 
                                  Sampling_unit == "lit" | 
                                  Sampling_unit == "pit" | 
                                  Sampling_unit == "permlit" |
                                  Sampling_unit == "permtrans" |
                                  Sampling_unit == "circular_transect" |
                                  Sampling_unit == "radial_transect"~ "Transect",
                                  Sampling_unit == "nd" ~ "ND",
                                  Sampling_unit == "stationary" ~ "Stationary",
                                  Sampling_unit == "expunit" ~ "Experimental unit",
                                  Sampling_unit == "sample" ~ "Sample",
                                  .default = "Other")
              )

# Change columns order
sampling_unit_tot <- cbind(sampling_unit_tot$Sampling_unit_macro,
                       sampling_unit_tot$Sampling_unit,
                       sampling_unit_tot$Count) 

colnames(sampling_unit_tot) <- c("Sampling_unit_macro","Sampling_unit","Count")

sampling_unit_tot <- as.data.frame(sampling_unit_tot)

sampling_unit_tot[,1:2] <- lapply(sampling_unit_tot[,1:2], as.factor)
sampling_unit_tot$Count <- as.numeric(sampling_unit_tot$Count)

sampling_unit_simplified <- aggregate(Count ~ Sampling_unit_macro, data = sampling_unit_tot, FUN = sum)

sampling_unit_simplified %<>%
  group_by(Sampling_unit_macro) %>%
  mutate(Percentage_macro= round(Count / sum(sampling_unit_simplified$Count) * 100,1))

sampling_unit_simplified
```

Build a treemap with simplified sampling unit

```{r Build a treemap with simplified sampling unit}
library(treemapify) 

RdYlBu <- RColorBrewer::brewer.pal(7, "RdYlBu")
RdYlBu

sampling_simplified_treemap <-
ggplot(sampling_unit_simplified,
       aes(area=Count,
           fill=Sampling_unit_macro,
           line = "black",
           label=paste0(Sampling_unit_macro, " \n ",Percentage_macro,"%"))
       )+ 
  treemapify::geom_treemap(layout = "squarified")+ 
  geom_treemap_text(place = "centre",
                    size = 20)+
    theme(plot.title = element_text(hjust = 0.5, size= 20),
          legend.position = "none")+
  scale_fill_manual(values= c("#FC8D59", "#FEE090","#FFFFBF", "#E0F3F8", "#D73027", "#91BFDB", "#4575B4"))
  scale_fill_brewer(palette = "RdYlBu")
  
sampling_unit_simplified %<>%
  mutate(case_when(Sampling_unit_macro == "Experimental unit" ~
          "Experimental \n unit", T ~ Sampling_unit_macro))
  
sampling_simplified_treemap <-
ggplot(sampling_unit_simplified,
       aes(area=Count,
           fill= reorder(Sampling_unit_macro, -Count),
           line = "black",
           label=paste0(Sampling_unit_macro, " \n ",Percentage_macro,"%"))
       )+ 
  treemapify::geom_treemap(layout = "squarified")+ 
  geom_treemap_text(place = "centre", fontface = "bold",
                    size = 20)+
    theme(plot.title = element_text(hjust = 0.5, size= 20),
          legend.position = "none")+
  scale_fill_manual(values=c("#01665E","#5AB4AC","#C7EAE5", "#8C510A","#D8B365","#F6E8C3","#F5F5F5")) # selected from "BrBG" palette of RBrewer

sampling_simplified_treemap
```

Export sampling_simplified_treemap

```{r Export sampling_simplified_treemap}
ggsave("sampling_simplified_treemap.png", 
       plot = sampling_simplified_treemap, 
       units = "in", width = 7, height = 6,
       path = here ("Extracted_data_analysis",
                    "polar",
                    "Output-Visualization"))
```

Export sampling_unit_tot table for supplementary materials

```{r Export sampling_unit_tot table for supplementary materials}
library(openxlsx)

openxlsx::write.xlsx(sampling_unit_tot,
           file = here("Extracted_data_analysis", "polar", "Output-Transformed_data",
                    "sampling_unit_tot.xlsx"))
```

## Quadrat extracted from transect

```{r}
print(paste("Number of publications where quadrat is the sampling unit or one of the sampling unit: ",
            nrow(quad_design <- subset(polar,grepl("quad", samplingUnit)
                                       )                                                                  )
            )
)

print(paste("Number of publications where quadrats were collected along transects: ", nrow(subset(quad_design, 
            grepl("transect\\(quadrat", meso_design) |
              grepl("transect\\(depth\\(quadrat", meso_design)
            )
     )
)
)
```
# DATA TYPE

```{r}
data_type <- count_results[["collectedDataType"]] 

data_type$Percentage <- round(data_type$Percentage, 1)

data_type
```

Lump uncommon combination into "Other"

```{r}
data_type_simplified <- data_type %>%
  rename(`Data type`= collectedDataType) %>%
  mutate(`Data type` = as.character(`Data type`)) %>%
  filter(Percentage > 2) %>%
  bind_rows(
    data_type %>%
      filter(Percentage <= 2) %>%
      summarise(
        `Data type` = "other",
        Count = sum(.data$Count),
        Percentage = sum(.data$Percentage)
      )
  )


data_type_simplified %<>%
       mutate(`Data type` = 
                case_when(`Data type` == 'photo'  ~ "Photo",
                          `Data type` == "video"~ "Video",
                          `Data type` == "other" ~ "Other",
                          `Data type` == "specimen" ~ "Specimen",
                          `Data type` == "photo+specimen" ~ "Photo + \n Specimen",
                          `Data type` == "photo+specimen+video+water" ~ "Photo + Video + \n Specimen \n Water",
                          `Data type` == "photo+video" ~ "Photo + \n Video",
                          `Data type` == "sediment+video" ~ "Sediment + \n Video",
                          `Data type` == "specimen+sediment+water" ~ "Specimen + \n Sediment + \n Water",
                          `Data type` == "specimen+video" ~ "Specimen + \n Video",
                          `Data type` == "specimen+video+water" ~ "Specimen + Water + \n Specimen",
                                  .default = "Other")
              )
```

Build a treemap with data type

```{r Build a treemap with data type}
library(treemapify) 
library(RColorBrewer)

data_type_treemap <-
ggplot(data_type_simplified,
       aes(area=Count,
           fill=`Data type`,
           line = "black",
           label=paste0(`Data type`, " \n ",Percentage,"%"))
       )+ 
  treemapify::geom_treemap(layout = "squarified")+ 
  geom_treemap_text(place = "centre", fontface = "bold",
                    size = 20)+
    theme(plot.title = element_text(hjust = 0.5, size= 20),
          legend.position = "none")+
    scale_fill_brewer(palette = "RdYlBu")

data_type_treemap
```

Export sampling_simplified_treemap

```{r Export data_type_treemap}
ggsave("data_type_treemap.png", 
       plot = data_type_treemap, 
       units = "in", width = 7, height = 6,
       path = here ("Extracted_data_analysis",                        "polar",                        "Output-Visualization"))
```


# HABIT

```{r}
habit <- as.data.frame(count_results["habit"]) 

# Rename columns 
habit %<>%
  rename(Habit = habit.habit,
         Count = habit.Count,
         Percentage = habit.Percentage)
```

Make the habit donut plot

```{r Make the habit donut plot}
habit %<>%
  mutate(ymax = cumsum(Count),
    ymin = c(0, head(ymax, n = -1)),
    mid = (ymin + ymax) / 2,
    angle = 90 - (mid / sum(Count)) * 360
  )

# Split and assign labels
habit_inside <- habit %>%
  mutate(label = paste0(round(Percentage,1), "%"))

habit_outside <- habit %>%
  mutate(label = Habit)

# Plot
habit_donut <- ggplot() +
  # Donut segments
  geom_rect(data = habit, 
            aes(ymin = ymin, ymax = ymax, xmin = 2.5, xmax = 4, fill = Habit),
            color = "black") +

  # Inside labels
  geom_text(data = habit_inside,
            aes(x = 3.2, y = mid, label = label),
            size = 5, color = "black",
            fontface = "bold") +

  # Outside labels with Habit name
  ggrepel::geom_text_repel(data = habit_outside,
                  aes(x = 4.3, y = mid, label = label),
                  size = 5.5,
                  nudge_x = 0.5,
                  direction = "y",
                  segment.color = "black",
                  hjust = 0,
                  box.padding = 0.3,
                  max.overlaps = Inf) +
  coord_polar(theta = "y") +
  xlim(0.2, 5) +
  scale_fill_brewer(palette = "Purples") +
  theme_void() +
  theme(legend.position = "none")+
  annotate(geom = 'text', x = 0.5, y = 0, 
           label = c("HABIT"), size = 8)

habit_donut
```

Export habit_donut

```{r Export habit_donut}
ggsave("habit_donut.png", 
       plot = habit_donut, 
       units = "in", width = 7, height = 6,
       path = here ("Extracted_data_analysis",                        "polar",                        "Output-Visualization"))
```

# SUBSTRATE

```{r substrate data frame}
substrate <- as.data.frame(count_results["meso_substrate"]) 

# Rename columns 
substrate %<>%
  rename(Substrate = meso_substrate.meso_substrate,
         Count = meso_substrate.Count,
         Percentage = meso_substrate.Percentage)

# Remove rows with nd
substrate <- substrate[!substrate$Substrate %in% "nd",]

# Merge substrate combination with less than 3% and recalculate Count and Percentages
substrate <- substrate %>%
  mutate(Substrate = if_else(Percentage < 3, "Other combinations", Substrate)) %>%
  group_by(Substrate) %>%
  select(-Percentage) %>%
  summarise(
    Count = sum(Count)) %>%
  mutate(Percentage = round(Count / sum(Count) * 100,1)
  )

substrate %<>%
  mutate(
    Substrate = case_when(Substrate == "hard" ~ "Hard",
                          Substrate == "soft"~ "Soft",
                          Substrate == "mixed" ~ paste0("Hard+","\n","Soft"),
                          Substrate == "artificial" ~ "Artificial",
                          Substrate == "Other combinations" ~ "Other combinations"))
```

Make the substrate donut plot

```{r substrate donut plot}
substrate %<>%
  mutate(
    ymax = cumsum(Count),
    ymin = c(0, head(ymax, n = -1)),
    mid = (ymin + ymax) / 2,
    angle = 90 - (mid / sum(Count)) * 360
  )


# Split and assign labels
substrate_inside <- substrate %>%
    filter(Percentage >= 5) %>%
  mutate(label = paste0(Percentage, "%"))

substrate_outside <- substrate %>%
  mutate(label = Substrate)

# Plot
substrate_donut <- ggplot() +
  # Donut segments
  geom_rect(data = substrate, 
            aes(ymin = ymin, ymax = ymax, xmin = 2.5, xmax = 4, fill = Substrate),
            color = "black") +

  # Inside labels
  geom_text(data = substrate_inside,
            aes(x = 3.2, y = mid, label = label),
            size = 5, color = "black",
            fontface = "bold") +

  # Outside labels with substrate name
  ggrepel::geom_text_repel(data = substrate_outside,
                  aes(x = 4.4, y = mid, label = label),
                  size = 5.5,
                  nudge_x = 0.3,
                  direction = "y",
                  segment.color = "black",
                  hjust = 0,
                  box.padding = 0.3,
                  max.overlaps = Inf) +
  coord_polar(theta = "y") +
  xlim(0.2, 5) +
  scale_fill_brewer(palette = "GnBu") +
  theme_void() +
  theme(legend.position = "none")+
  annotate(geom = 'text', x = 0.5, y = 0, 
           label = c("SUBSTRATE"), size = 8)

substrate_donut
```
Export the substrate donut plot

```{r Export the substrate donut plot}
ggsave("substrate_donut.png", 
       plot = substrate_donut,
       units = "in", width = 7, height = 6,
       path = here ("Extracted_data_analysis",                        "polar",                        "Output-Visualization"))
```

# BENTHIC POSITION

Create a data frame with benthic position

```{r position data frame}
position <- as.data.frame(count_results["environmentalPosition"]) 

# Rename columns 
position %<>%
  rename(Position = environmentalPosition.environmentalPosition,
         Count = environmentalPosition.Count,
         Percentage = environmentalPosition.Percentage)

# Merge position combination with less than 3% and recalculate Count and Percentages
position <- position %>%
  mutate(Position = if_else(Percentage < 4, "Other combinations", Position)) %>%
  group_by(Position) %>%
  select(-Percentage) %>%
  summarise(
    Count = sum(Count)) %>%
  mutate(Percentage = round(Count / sum(Count) * 100,1)
  )
```

Make the benthic position donut plot

```{r benthic position donut plot}
position %<>%
  mutate(
    Position = case_when(Position == "epi" ~ "Epi",
                          Position == "endo"~ "Endo",
                          Position == "dem+epi" ~ "Hyper + \n Epi",
                          Position == "endo+epi" ~ "Endo + \n Epi",
                          Position == "Other combinations" ~ "Other combinations"),
    ymax = cumsum(Count),
    ymin = c(0, head(ymax, n = -1)),
    mid = (ymin + ymax) / 2,
    angle = 90 - (mid / sum(Count)) * 360
  )


# Split and assign labels
position_inside <- position %>%
  filter(Percentage > 3) %>%
  mutate(label = paste0(Percentage, "%"))

position_outside <- position %>%
  mutate(label = Position)

# Plot
benthic_position_donut <- ggplot() +
  # Donut segments
  geom_rect(data = position, 
            aes(ymin = ymin, ymax = ymax, xmin = 2.5, xmax = 4, fill = Position),
            color = "black") +

  # Inside labels
  geom_text(data = position_inside,
            aes(x = 3.2, y = mid, label = label),
            size = 5, color = "black",
            fontface = "bold") +

  # Outside labels with Position name
  ggrepel::geom_text_repel(data = position_outside,
                  aes(x = 4.2, y = mid, label = label),
                  size = 5.5,
                  nudge_x = 0.3,
                  direction = "y",
                  segment.color = "black",
                  hjust = 0,
                  box.padding = 0.3,
                  max.overlaps = Inf) +
  coord_polar(theta = "y") +
  xlim(0.2, 5) +
  scale_fill_brewer(palette = "OrRd") +
  theme_void() +
  theme(legend.position = "none")+
  annotate(geom = 'text', x = 0.5, y = 0, 
           label = c("BENTHIC \n POSITION"), size = 8)

benthic_position_donut
```

Export the benthic_position donut plot

```{r Export the benthic_position donut plot}
ggsave("benthic_position_donut.png", 
       plot = benthic_position_donut, 
       units = "in", width = 7, height = 6,
       path = here ("Extracted_data_analysis",                        "polar",                        "Output-Visualization"))
```

# TARGET 

## Target level 

```{r}
target_level <- as.data.frame(count_results["taxonomicTargetLevel"]) 

# Rename columns 
target_level %<>%
  rename(Level = taxonomicTargetLevel.taxonomicTargetLevel,
         Count = taxonomicTargetLevel.Count,
         Percentage = taxonomicTargetLevel.Percentage)

# Remove rows with nd
target_level <- target_level[!target_level$Level %in% "species,species",]
target_level <- target_level[!target_level$Level %in% ",species",]

# Merge habit combination with less than 3% and recalculate Count and Percentages
target_level <- target_level %>%
  mutate(Level = if_else(Percentage < 3, "Other combinations", Level)) %>%
  group_by(Level) %>%
  select(-Percentage) %>%
  summarise(
    Count = sum(Count)) %>%
  mutate(Percentage = round(Count / sum(Count) * 100,1)
  )
```

Treemap with the taxonomic target level

```{r Treemap with the taxonomic target level}
target_level_labels <- target_level %<>%
  mutate(
    Level = case_when(Level == "comm" ~ "Community",
                      Level == "group"~ "Group",
                      Level == "single" ~ "Single species",
                      Level == "species" ~ ">1 species",
                      Level == "comm+group" ~ "Community+Group",
                      Level == "comm+species" ~ "Community+Species",
                      Level == "group+species" ~ "Group+\nSpecies", 
                      Level == "Other combinations" ~ "Other")
    )

target_level_treemap <- ggplot(target_level_labels,
       aes(area=Count,
           fill=Level,
           line = "black",
           label=paste0(Level, 
                        " \n ",Percentage,"%"))
       )+ 
  treemapify::geom_treemap(layout = "squarified")+ 
  treemapify::geom_treemap_text(place = "centre",
                    size = 20)+
    theme(plot.title = element_text(hjust = 0.5, size= 20),
          legend.position = "none")+
  scale_fill_manual(values= c("#A6D854" ,"#66C2A5","#E78AC3", "#8DA0CB","#FFD92F","#B3B3B3", "#E5C494", "#FC8D62"))

target_level_treemap
```

```{r}
ggsave("target_level_treemap.png",
       plot = target_level_treemap,
       units = "in", width = 9, height = 6,
       path = here ("Extracted_data_analysis",                        "polar",                        "Output-Visualization"))
```


## Detected phyla in community studies

```{r}
detected <- polar %>% 
    filter(str_detect(taxonomicTargetLevel,"comm")) %>%
  select(File, 
         detected_meso_cnidaria, 
         detected_meso_sponges, 
         detected_meso_seagrass, 
         detected_meso_algae, 
         detected_meso_mollusca, 
         detected_meso_chondrichthyes, 
         detected_meso_fish, 
         detected_meso_crustacea, 
         detected_meso_echino, 
         detected_meso_other_taxa)
```


```{r}
detected_lvls <- 
  unique(
    unlist(
      str_split(detected$detected_meso_other_taxa, "\\+")
                               )
                        )
detected_lvls<- detected_lvls[! detected_lvls %in% c("",NA,"notapp",",","nd")]
detected[,detected_lvls] <- sapply(detected_lvls, 
                                      function(y){
                                        str_count(detected$detected_meso_other_taxa, y)
                                        }
                                      )

rm(detected_lvls)

detected %<>%
  mutate(across(everything(), as.character)) %>%
  replace(is.na(.), "0") %>%
  select(-detected_meso_other_taxa) 

detected <- detected %>%
  mutate(across(-File, ~ case_when(
    .x %in% c("x", "x,", "xx") ~ "1",
    .x == "," ~ "0",
    TRUE ~ .x
  ))) %>%
  mutate(across(-File, as.numeric))  
```

Calculate number of publications per taxonomic group 

```{r}
detected_tot <- 
  detected %>%
  summarise(across(where(is.numeric), 
                   list(sum)
                   )
            )

detected_tot  <- reshape2::melt(detected_tot, id.vars = NULL) 

colnames(detected_tot) <- c("Taxa", "Count")

detected_tot %<>%
        mutate_at("Taxa", 
                  stringr::str_replace, "_1", "") %>%
        mutate_at("Taxa", 
                  stringr::str_replace, "detected_meso_", "")

    
detected_tot %<>%
      mutate(Percentage= round(Count / nrow(detected) * 100,1))  

rm(detected)
```

Remove absent taxonomic groups

```{r Merge poorly represented taxonomic groups into "Other"}
detected_tot_collapsed <- subset(detected_tot, detected_tot$Count>0)
```

Create a bar chart with detected taxonomic groups

```{r detected_bar}

detected_bar <- ggplot(detected_tot_collapsed, aes(x = reorder(Taxa, Count), y = Count, fill= Taxa)) +
  geom_bar(stat = "identity") +
  geom_text(
    data = detected_tot_collapsed %>% filter(Percentage > 11),
    aes(label = paste0(round(Percentage, 0), "%")),
    position = position_stack(vjust = 0.5),
    color = "black",
    size = 3.5
  )+
  labs(y = "# Publications at the community level", x = "Detected taxonomic group") +
  theme_minimal() +
  coord_flip()+  # optional: flips for horizontal bars
  theme(legend.position ="none") +
  scale_x_discrete(labels = rev(c("Echinodermata",
                                "Porifera",
                                "Cnidaria",
                                "Mollusca",
                                "Crustacea",
                                "Annelida",
                                "Tunicata",
                                "Bryozoa",
                                "Osteichthyes",
                                "Pycnogonida",
                            "macroalgae*",
                            "Brachiopoda",
                            "Nemertea",
                            "Nematoda",
                            "Hemichoradata",
                            "Foraminifera")))

detected_palette <- c("algae" ="darkseagreen3",
                      "cnidaria" = "mediumvioletred",
                      "crustacea" = "indianred1",
                      "sponges" = "khaki1",
                      "mollusca" = "darkorange3",
                      "fish" = "paleturquoise2",
                      "echino" = "royalblue4",
                      "Tunicata" = "mediumpurple4", 
                      "Hemichordata" = "gray69",
                      "Brachiopoda" = "gray69",
                      "Nemertea" = "gray69",
                      "Pycnogonida" = "gray69",
                      "Nematoda" = "gray69",
                      "Foraminifera"   =    "gray69",
                      "Bryozoa" = "goldenrod1",
                      "Annelida" = "burlywood4")



detected_bar + 
  scale_fill_manual(values = detected_palette,
    limits = names(detected_palette))+
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16)
        )
  
```

```{r}
ggsave("detected_bar.png", 
       units = "in", width = 7, height = 6,
       path = here ("Extracted_data_analysis",                        "polar",                        "Output-Visualization"))
```

## Target taxonomic groups not at the community level

```{r create group data frame}
group <- polar %>% 
    filter(taxonomicTargetLevel != "comm") %>%
  select(File, 
         detected_meso_cnidaria, 
         detected_meso_sponges, 
         detected_meso_seagrass, 
         detected_meso_algae, 
         detected_meso_mollusca, 
         detected_meso_chondrichthyes, 
         detected_meso_fish, 
         detected_meso_crustacea, 
         detected_meso_echino, 
         detected_meso_other_taxa,
         targetTaxa)
```


```{r}
group_lvls <- 
  unique(
    unlist(
      str_split(group$detected_meso_other_taxa, "\\+")
                               )
                        )
group_lvls<- group_lvls[! group_lvls %in% c("",NA,"notapp",",","nd")]
group[,group_lvls] <- sapply(group_lvls, 
                                      function(y){
                                        str_count(group$detected_meso_other_taxa, y)
                                        }
                                      )

rm(group_lvls)
```


```{r}
group <- group %>%
  mutate(across(everything(), as.character)) %>%
  replace(is.na(.), "0") %>%
  select(-detected_meso_other_taxa,-targetTaxa) %>%
  mutate(across(-File, ~ case_when(
    .x %in% c("x", "x,", "xx") ~ "1",
    .x == "," ~ "0",
    TRUE ~ .x
  ))) %>%
  mutate(across(-File, as.numeric))
```


```{r}
group_tot <- 
  group %>%
  summarise(across(where(is.numeric), 
                   list(sum)
                   )
            )

group_tot  <- reshape2::melt(group_tot, id.vars = NULL) 

colnames(group_tot) <- c("Taxa", "Count")

group_tot %<>%
        mutate_at("Taxa", 
                  stringr::str_replace, "_1", "") %>%
        mutate_at("Taxa", 
                  stringr::str_replace, "detected_meso_", "")

    
group_tot %<>%
      mutate(Percentage= round(Count / nrow(group) * 100,1))  
```

```{r}
group_tot_collapsed <- group_tot %>%
  filter(Percentage > 0 ) %>%
  group_by(Taxa) %>%
  summarise(
    Count = sum(Count),
    Percentage = sum(Percentage),
    .groups = "drop"
  )
```


```{r}
group_tot_collapsed %<>%
  mutate(Taxa = case_when(Taxa == 'echino' ~ "Echinodermata",
                          Taxa == 'fish' ~ "Osteichthyes",
                          Taxa == 'mollusca' ~ "Mollusca",
                          Taxa == 'algae' ~ "macroalgae*",
                          Taxa == "sponges" ~ "Porifera",
                          Taxa == "crustacea" ~ "Crustacea",
                          Taxa == "cnidaria" ~ "Cnidaria",
                          Taxa == "algae" ~ "algae*",
         .default = Taxa))
```


```{r}
group_bar <- ggplot(group_tot_collapsed, aes(x = reorder(Taxa, Count), y = Count, fill= Taxa)) +
  geom_bar(stat = "identity") +
  geom_text(
    data = group_tot_collapsed %>% filter(Percentage > 10),
    aes(label = paste0(round(Percentage, 0), "%")),
    position = position_stack(vjust = 0.5),
    color = "black",
    size = 3.5
  )+
  labs(y = "N of publications not at community level", x = "Taxonomic group") +
  theme_minimal() +
  coord_flip()+  # optional: flips for horizontal bars
  theme(legend.position ="none")

group_palette <- c("macroalgae*" ="darkseagreen3",
                      "Cnidaria" = "mediumvioletred",
                      "Crustacea" = "indianred1",
                      "Porifera" = "khaki1",
                      "Mollusca" = "darkorange3",
                      "Osteichthyes" = "paleturquoise2",
                      "Echinodermata" = "royalblue4",
                      "Tunicata" = "mediumpurple4", 
                      "Chondrichthyes" = "gray69",
                      "Brachiopoda" =  "gray69",
                      "Pycnogonida" =  "gray69",
                      "Bryozoa" = "goldenrod1",
                      "Annelida" = "burlywood4")

group_bar <- group_bar + 
  scale_fill_manual(values = group_palette,
    limits = names(group_palette))+
    theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16)
        )

group_bar
```
Export bar plot 

```{r export group_bar.png}
ggsave("group_bar.png", 
       plot = group_bar,
       units = "in", width = 7, height = 6,
       path = here ("Extracted_data_analysis",                        "polar",                        "Output-Visualization"))
```

## New species described per phyla

```{r}
new_spec <- polar %>%
  filter(str_detect(taxonomic_publication, "taxa")) 

new_spec %<>% 
  select(File, 
         detected_meso_cnidaria, 
         detected_meso_sponges, 
         detected_meso_seagrass, 
         detected_meso_algae, 
         detected_meso_mollusca, 
         detected_meso_chondrichthyes, 
         detected_meso_fish, 
         detected_meso_crustacea, 
         detected_meso_echino, 
         detected_meso_other_taxa)
```


# DATA CROSSING

```{r}
table(polar$meso_protection)
```

## VISUALLY SURVEYED SURFACE

```{r}
polar$visuallySurveyedSurface <- as.factor(polar$visuallySurveyedSurface)
summary(subset(polar, is.na(polar$taxonomic_publication))$visuallySurveyedSurface)
```

## Observation

```{r}
print(paste("Number of manipulative experimental studies:",nrow(subset(polar, grepl("exp", polar$observationalORexperimental)))))

print(paste("Number of manipulative experimental studies in the field of restoration:", nrow(subset(polar, grepl("exp", polar$observationalORexperimental) & polar$field_rest != 0)
                                                                                            )
            )
      )
```

## Temp replicates & exp

```{r # Publications that included temporal replicates}
print(paste("# Publications that included temporal replicates:", nrow(subset(polar, grepl("yes", temporalReplicates)))))
```

```{r  # Publications relying on a manipulative experiment that included temporal replicates}
print(paste("# Publications relying on a manipulative experiment that included temporal replicates:", nrow(subset(polar, (grepl("exp", polar$observationalORexperimental) & grepl("yes", temporalReplicates))))))
```

## ROV & sample

```{r}
rov <- subset(polar, grepl("rov", meso_samplingPlatform) )

rov_specimen <- subset(rov,  grepl("specimen", collectedDataType) )

print(paste("Percentage of publications relying solely on ROV where specimens were collected:", round(nrow(rov_specimen)/nrow(rov)*100,1),"%"))

rm(rov, rov_specimen)
```

## Molecular techniques in taxonomic and non-taxonomic studies

```{r Molecular techniques in non-taxonomic studies}
polar$taxonomic_publication <- as.factor(polar$taxonomic_publication )

molecular <- polar %>%
  filter(molecularAnalyses != "no" & 
           molecularAnalyses != "notapp" & 
           molecularAnalyses != "nd")

print(paste("# Publications that adopted molecular techniques: ",nrow(molecular)))

molecular_notaxa <- molecular %>%
  filter(is.na(taxonomic_publication))

print(paste("# Publications that adopted molecular techniques not for taxonomy: ",nrow(molecular_notaxa)))

rm(molecular, molecular_notaxa)
```



